$\pi$ with leftovers: \\ a mechanisation in Agda

### 2020-12-14 paper/paper.tex
Remove the index from process termination
Adapt to LLNCS style

\documentclass[]{lipics-v2019}

@@

% Do not use italics in definitions and theorems
\theoremstyle{definition}
\newtheorem{nidefinition}{Definition}
\newtheorem{nitheorem}{Theorem}
\newtheorem{nilemma}{Lemma}
% \newtheorem{example}{Example}
\renewcommand{\nidefinitionautorefname}{Definition}%
\renewcommand{\nitheoremautorefname}{Theorem}%
\renewcommand{\nilemmaautorefname}{Lemma}%
\renewcommand{\sectionautorefname}{\S}%
\renewcommand{\subsectionautorefname}{\S}%
% \renewcommand{\exampleautorefname}{Example}

@@

% Functions
\newcommand{\subst}[3]{#1 \, \func{[} \, #3 \, \func{\mapsto} \,#2 \, \func{]}}
\newcommand{\op}[3]{#1 \, \func{\coloneqq} \, #2 \, \func{\cdot} \, #3}
\newcommand{\opsquared}[3]{#1 \, \func{\coloneqq} \, #2 \, \func{\cdot^2} \, #3}
\newcommand{\opctx}[3]{#1 \, \func{\coloneqq} \, #2 \, \func{\otimes} \, #3}
\newcommand{\zero}{\func{0\cdot}}
\newcommand{\one}{\func{1\cdot}}
\newcommand{\li}{\func{\ell_i}}
\newcommand{\lo}{\func{\ell_o}}
\newcommand{\lz}{\func{\ell_{\varnothing}}}
\newcommand{\lio}{\func{\ell_{\#}}}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\title{$\pi$ with leftovers: \\ a mechanisation in Agda}
\titlerunning{$\pi$ with leftovers: a mechanisation in Agda}
\author{Uma Zalakain}{University of Glasgow, Scotland}{u.zalakain.1@research.gla.ac.uk}{https://orcid.org/0000-0002-3268-9338}{}
\author{Ornela Dardha}{University of Glasgow, Scotland}{ornela.dardha@glasgow.ac.uk}{https://orcid.org/0000-0001-9927-7875}{}
\keywords{pi-calculus, linear types, leftover typing, concurrency, mechanisation, Agda}
\Copyright{U. Zalakain, and O. Dardha}
\ccsdesc[300]{Theory of computation~Process calculi}
% \supplement{\url{https://github.com/umazalakain/typing-linear-pi}}
\begin{CCSXML}
<ccs2012>
<concept>
<concept_id>10003752.10003753.10003761.10003764</concept_id>
<concept_desc>Theory of computation~Process calculi</concept_desc>
<concept_significance>300</concept_significance>
</concept>
</ccs2012>
\end{CCSXML}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

\acknowledgements{We want to thank Erika Kreuter, Wen Kokke, James Wood, Guillaume Allais, Bob Atkey, and Conor McBride for their thoughts, patience, work, education and camaraderie.}

\funding{The authors would also like to thank the anonymous referees for their valuable comments and helpful suggestions. This work is supported by the UK EPSRC grant EP/K034413/1 ``From Data Types to Session Types: A Basis for Concurrency and Distribution'' (ABCD), and by the EU HORIZON 2020 MSCA RISE project 778233 ``Behavioural Application Program Interfaces'' (BehAPI).}

@@

\begin{abstract}
  The \picalc{} is a computational model for communication and concurrency.
  The \emph{linear} \picalc{} is a typed version of the \picalc{} where channels must be used exactly once.
  It is an underlying theoretical and practical framework on top of which more advanced types and theories can be built, such as \emph{session types} \cite{H93,THK94,HVK98}.

  We present an untyped syntax and semantics for the \picalc{}, on top of which we provide a resource-aware type system parametrised by a set of multiplicity algebras.
  This allows the user to mix \emph{linear}, \emph{graded} and \emph{shared} types, under the same unified framework.
  We use leftover typing \cite{Allais2018a} to encode our typing rules, thus avoiding the ubiquitous context splitting proofs common when dealing with linear type systems.
  We provide a framing theorem, generalise the weakening and strengthening theorems, and use them to prove subject congruence.
  We show that the type system is stable under substitution and prove subject reduction.
%
  Our formalisation is fully mechanised in Agda.
\end{abstract}

@@

In this paper we make the following contributions:
\begin{enumerate}
  \item We use \textbf{type-level {de Bruijn indices} \cite{deBruijn1972, Dybjer1994} to introduce a syntax for \picalc{} processes that is well-scoped by construction}:
    Every free variable is accounted for in the type of the process that uses it (\autoref{syntax}).
    \todo{Do Ciccone \& Padovani also use type level de Bruijn indices?}

@@

\subsection{Summary}
\todo{shorten}
\begin{enumerate}
\item \textbf{Formalisation of the syntax and the semantics of the \picalc{}}:
 \begin{itemize}
   \item \textbf{Syntax}: we use type level {de Bruijn indices} \cite{deBruijn1972, Dybjer1994} to introduce a syntax of \picalc{} processes that is \emph{well scoped by construction}: every free variable is accounted for in the type of the process that uses it (\autoref{syntax}).
   For convenience and readability, the user can convert from and to a syntax using channel names (\autoref{from_to_deBruijn}).
   These conversion functions are shown to be inverses of one another up to $\alpha$-conversion.

@@

  \begin{equation*}
    \begin{aligned}
      \Process_n ::=& \; \PO_n               &&\text{(inaction)}   \\ 
      |& \; \new{} \Process_{\suc n}         &&\text{(restriction)}          \\ 
      |& \; \comp{\Process_n}{\Process_n}    &&\text{(parallel)} \\ 
      |& \; \recv{\Var_n}{}\Process_{\suc n} &&\text{(input)}                \\ 
      |& \; \send{\Var_n}{\Var_n}\Process_n  &&\text{(output)}                  
    \end{aligned}
  \end{equation*}
  \caption{Well-scoped grammar with type-level de Bruijn indices.}
  \label{fig:syntax}
\end{figure}

@@

\begin{nidefinition}
  \label{def:unused}
  A variable $i$ is unused in $P$ ($\Unused_i \; P$) if it appears in none of its inputs and none of its outputs.
  \todo{should we have this as a figure instead?}

@@

  The type $i \type{\not\equiv} x$ unfolds to the negation of propositional equality on \textsc{Var}, i.e. $i \type{\equiv} x \to \type{\bot}$.
\end{nidefinition}

@@

\begin{nilemma}
  \label{lm:substitution-unused}
  For every variables $i$ and $j$, if $i \type{\not\equiv} j$ then $\Unused_i \allowbreak (\subst{P}{j}{i})$.
\end{nilemma}
\begin{proof}
  By structural induction on \textsc{Process} and \textsc{Var}.
\end{proof}

@@

\begin{nidefinition}[Usage algebra]
  A \emph{usage algebra} is a ternary relation $\op{x}{y}{z}$ that is \emph{partial} (as not any two multiplicities can be combined), \emph{deterministic} and \emph{cancellative} (to aid equational reasoning), \emph{associative} and \emph{commutative} (following directly from subject congruence for parallel composition), and in which the leftovers can be \emph{computed} (as to automatically update the usage context every time input and output occurs).
  It has a \emph{neutral element} $\zero$ that is absorbed on either side, and that is also \emph{minimal} (so that new resources cannot arbitrarily spring into life).
  It has an element $\one$ that is used to count inputs and outputs.
  \autoref{fig:multiplicities} defines such an algebra as a record $\Algebra_C$ on a carrier $C$.

  \begin{figure}[h]
  \begin{equation*}
  \begin{aligned}
    &\zero                   &:{} &                 &        & C \\
    &\one                    &:{} &                 &        & C \\
    &\op{\_}{\_}{\_}         &:{} &                 &        & C \to C \to C \to \Set \\
    &\func{\cdot-compute^r}  &:{} &\forall x y       & \to \; & \type{DEC} \; (\type{\exists} z  \; (\op{x}{y}{z})) \\
    &\func{\cdot-unique}     &:{} &\forall x x' y z  & \to \; & \op{x'}{y}{z} \to \op{x}{y}{z} \to x' \equiv x \\
    &\func{\cdot-unique^l}   &:{} &\forall x y y' z  & \to \; & \op{x}{y'}{z} \to \op{x}{y}{z} \to y' \equiv y \\
    &\func{\cdot-min^l}      &:{} &\forall y z        & \to \; & \op{\zero}{y}{z} \to y \equiv \zero \\
    &\func{\cdot-id^l}       &:{} &\forall x         & \to \; & \op{x}{\zero}{x} \\
    &\func{\cdot-comm}       &:{} &\forall x y z     & \to \; & \op{x}{y}{z} \to \op{x}{z}{y} \\
    &\func{\cdot-assoc}      &:{} &\forall x y z u v & \to \; & \op{x}{y}{z} \to \op{y}{u}{v} \to \\
    &                        &    &                  &        & \type{\exists} w  \; (\op{x}{u}{w} \times \op{w}{v}{z})
  \end{aligned}
  \end{equation*}
  \caption{
    Usage algebra $\Algebra_C$ on a carrier $C$.
    We use $\forall$ for universal quantification.
    The dependent product $\type{\exists}$ uses the value of its first argument in the type of its second.
    The type $\type{DEC} \; P$ is a witness of either $P$ or $P \to \type{\bot}$, where $\type{\bot}$ is the empty type with no constructors.
  }
  \label{fig:multiplicities}
  \end{figure}
\end{nidefinition}

@@

We use indexed sets of usage algebras to allow several usage algebras to coexist in our type system with leftovers (\autoref{leftover-typing}).
\begin{nidefinition}[Indexed set of usage algebras]
  An \emph{indexed set of usage algebras} is a type $\Idx$ of indices that is nonempty ($\type{\exists IDX}$) together with an interpretation $\Usage$ of indices into types, and an interpretation $\type{ALGEBRAS}$ of indices into usage algebras of the corresponding type (\autoref{fig:indexed-multiplicities}).
  \begin{figure}[h]
    \begin{equation*}
      \begin{aligned}
        &\Idx               &:{} &\Set \\
        &\type{\exists IDX} &:{} &\Idx \\
        &\Usage             &:{} &\Idx \to \Set \\
        &\type{ALGEBRAS}    &:{} &(idx : \Idx) \to \Algebra_{\Usage_{idx}}
      \end{aligned}
    \end{equation*}
    \caption{Indexed set of usage algebras.}
    \label{fig:indexed-multiplicities}
  \end{figure}
\end{nidefinition}

@@

\begin{nidefinition}[\textsc{Type} and \textsc{PreCtx}: types and typing contexts]
  A \emph{type} is either a unit type ($\unit$), a base type ($\base{n}$) or a channel type ($\channel{t}{x}$) (first two rows of \autoref{fig:prectx}).

@@

A \emph{typing context} (last two rows of \autoref{fig:prectx}) is a length-indexed list of types, and is either empty $[]$ or the result of appending a type $t$ to an existing context $\gamma$.
\end{nidefinition}

\begin{nidefinition}[\textsc{Ctx}: usage contexts]
  A \emph{usage context} is a context $\Ctx_{idxs}$ of pairs of usage annotations that is indexed by a length-indexed context of indices $\Idxs_n$ (\autoref{fig:ctx}).

@@

    \inferrule
        {\Gamma : \Ctx_{idxs} \\ x : \Usage_{idx} ^{\func{2}}}
        {\Gamma \comma x : \Ctx_{idxs \comma idx}}
  \end{mathpar}
  \caption{Length-indexed context of carrier indices with a context of usage annotations on top.}
  \label{fig:ctx}
  \end{figure}
\end{nidefinition}

@@

\begin{nidefinition}[\textsc{VarRef}: typing variable references]
  \label{def:varref}

@@

  The base case $\constr{0}$ splits the usage annotation $x$ of type $\Usage_{idx}$ into $y$ and $z$ (the leftovers).
  Note that the remaining context $\Gamma$ is preserved unused as a leftover.
  This splitting $\opsquared{x}{y}{z}$ is as per the usage algebra provided by the developer for the index $idx$.
  We use the \emph{computability} of the monoidal relation to alleviate the user from the proof burden $\opsquared{x}{y}{z}$.
  The inductive case $\suc$ appends the type $t'$ to the typing context, and the usage annotation $x'$ to both the input and output usage contexts. 
\end{nidefinition}

@@

\begin{nidefinition}[\textsc{Types}: typing processes]
  \label{def:types}

@@

  The inaction process in rule $\PO$ does not change usage annotations.
  The scope restriction in rule $\new$ expects three arguments: the type $t$ of data being transmitted; the usage annotation $x$ of what is being transmitted; and the multiplicity $y$ given to the channel itself.
  This multiplicity $y$ is used for both input and output, so that they are balanced.
  The continuation process $P$ is provided with the new channel with usage annotation $y \comma y$, which it must completely exhaust.
%
  The input process in rule $\recv{}{}$ requires a channel $chan_i$ at index $i$ with usage $\li$ available, such that data with type $t$ and usage $x$ can be sent over it.
  Note that the index $i$ is used in the syntax of the typed process.
  We use the leftovers $\Xi$ to type the continuation process, which is also provided with the received element --- of type $t$ and multiplicity $x$ --- at index $\constr{0}$.
  The received element $x$ must be completely exhausted by the continuation process.
%
  Similarly to input, the output process in rule $\send{}{}$ requires a channel $chan_i$ at index $i$ with usage $\lo$ available, such that data with type $t$ and usage $x$ can be sent over it.
  We use the leftover context $\Delta$ to type the transmitted data, which needs an element $loc_j$ at index $j$ with type $t$ and usage $x$, as per the type of the channel $chan_i$.
  The leftovers $\Xi$ are used to type the continuation process.
  Note that both indices $i$ and $j$ are used in the syntax of the typed process.
%
  Parallel composition in rule $\comp{}{}$ uses the leftovers of the left-hand process to type the right-hand process.
  By keeping track in the typing derivation of $P$ of the resources $P$ uses, we can use them to type $Q$ and save the user from having to provide a top-down proof of context split.
\end{nidefinition}

@@

\begin{nitheorem}[Framing]
  \label{thm:framing}
  Let $P$ be well typed in $\types{\gamma}{\Gamma_l}{P}{\Xi_l}$.
  Let $\Delta$ be such that $\opctx{\Gamma_l}{\Delta}{\Xi_l}$.
  Let $\Gamma_r$ and $\Xi_r$ be arbitrary contexts such that $\opctx{\Gamma_r}{\Delta}{\Xi_r}$.
  Then $\types{\gamma}{\Gamma_r}{P}{\Xi_r}$.
\end{nitheorem}

\paragraph*{Weakening}
This property states that inserting a new variable into the context preserves the well-typedness of a process as long as the usage annotation of the inserted variable is preserved as a leftover.
Let $\func{ins}_i$ insert an element into a context at position $i$ --- for simplicity, we use it both to insert types into typing contexts and usage annotations into usage contexts.
\begin{nitheorem}[Weakening]
  \label{thm:weakening}
  Let $P$ be well typed in $\types{\gamma}{\Gamma}{P}{\Xi}$.
  Then, lifting every variable greater than or equal to $i$ in $P$ is well typed in
  $\types{\func{ins}_i \; t \; \gamma}{\func{ins}_i \; x \; \Gamma}{\func{lift}_i \; P}{\func{ins}_i \; x \; \Xi}$.
\end{nitheorem}

\paragraph*{Strengthening}
This property states that removing an unused variable preserves the well-typedness of a process.
Let $\func{del}_i$ delete the element at position $i$ from a context --- for simplicity, we use it both to delete types from typing contexts and usage annotations from usage contexts.
\begin{nitheorem}[Strengthening]
  \label{thm:strengthening} 
  Let $P$ be well typed in $\types{\gamma}{\Gamma}{P}{\Xi}$.
  Let $i$ be a variable not in $P$, such that $uP \; : \; \Unused_i \; P$.
  Then lowering every variable greater than $i$ in $P$ is well typed in $\types{\func{del}_i \; \gamma}{\func{del}_i \; \Gamma}{\func{lower}_i \; P \; uP}{\func{del}_i \; \Xi}$.
\end{nitheorem}

\paragraph*{Exchange}
This property states that the exchange of two variables preserves the well-typedness of a process.
We extend $\func{exchange}_i$  introduced in \autoref{structural-congruence} to exchange types in typing contexts and usage annotations in usage contexts.
\begin{nitheorem}[Exchange]
  \label{thm:exchange}
  Let $P$ be well typed in $\types{\gamma}{\Gamma}{P}{\Xi}$.
  Then, $\types{\func{exchange}_i \; \gamma}{\func{exchange}_i \; \Gamma}{\allowbreak \func{exchange}_i \; P}{\func{exchange}_i \; \Xi}$.
\end{nitheorem}

@@

\begin{nilemma}
  \label{lm:lower-lift}
  The function $\func{lower}_i \; P \; uP$ has an inverse $\func{lift}_i \; P$ that increments every $\textsc{Var}$ greater than or equal to $i$, such that $\func{lift}_i \; (\func{lower}_i \; P \; uP) \equiv P$.
\end{nilemma}
\begin{proof}
  By structural induction on \textsc{Process} and \textsc{Var}.
\end{proof}

\begin{nilemma}
  \label{lm:exchange-exchange}
  The function $\func{exchange}_i \; P$ is its own inverse: $\func{exchange}_i \; (\func{exchange}_i \; P) \equiv P$.
\end{nilemma}
\begin{proof}
  By structural induction on \textsc{Process} and \textsc{Var}.
\end{proof}

\begin{nilemma}
  \label{lm:types-unused}
  For all well-typed processes $\types{\gamma}{\Gamma}{P}{\Xi}$, if the variable $i$ is unused within $P$, then $\Gamma$ at $i$ is equivalent to $\Xi$ at $i$.
\end{nilemma}
\begin{proof}
  By induction on \textsc{Process} and \textsc{Var}.
\end{proof}

@@

\begin{nitheorem}[Subject congruence]
  \label{thm:subject-congruence}
  If $P \eq{} Q$ and $\types{\gamma}{\Gamma}{P}{\Xi}$, then $\types{\gamma}{\Gamma}{Q}{\Xi}$.
\end{nitheorem}

>>>>

\documentclass[]{llncs}

@@

% Do not use italics in definitions and theorems
% \theoremstyle{definition}
% \newtheorem{definition}{Definition}
% \newtheorem{theorem}{Theorem}
% \newtheorem{lemma}{Lemma}
% % \newtheorem{example}{Example}
% \renewcommand{\definitionautorefname}{Definition}%
% \renewcommand{\theoremautorefname}{Theorem}%
% \renewcommand{\lemmaautorefname}{Lemma}%
% \renewcommand{\sectionautorefname}{\S}%
% \renewcommand{\subsectionautorefname}{\S}%
% \renewcommand{\exampleautorefname}{Example}

@@

% Functions
\newcommand{\subst}[3]{#1 \, \func{[} \, #3 \, \func{\mapsto} \,#2 \, \func{]}}
\newcommand{\op}[3]{#1 \, \func{\coloneqq} \, #2 \, \func{\cdot} \, #3}
\newcommand{\opsquared}[3]{#1 \, \func{\coloneqq} \, #2 \, \func{\cdot^2} \, #3}
\newcommand{\opctx}[3]{#1 \, \func{\coloneqq} \, #2 \, \func{\otimes} \, #3}
\newcommand{\zero}{\func{0\cdot}}
\newcommand{\one}{\func{1\cdot}}
\newcommand{\li}{\func{\ell_i}}
\newcommand{\lo}{\func{\ell_o}}
\newcommand{\lz}{\func{\ell_{\varepsilon}}}
\newcommand{\lio}{\func{\ell_{\#}}}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\title{
  $\pi$ with leftovers: \\ a mechanisation in Agda
  \thanks{The authors would also like to thank the anonymous referees for their valuable comments and helpful suggestions. This work is supported by the UK EPSRC grant EP/K034413/1 ``From Data Types to Session Types: A Basis for Concurrency and Distribution'' (ABCD), and by the EU HORIZON 2020 MSCA RISE project 778233 ``Behavioural Application Program Interfaces'' (BehAPI).}
}
\titlerunning{$\pi$ with leftovers: a mechanisation in Agda}
\author{
  Uma Zalakain\inst{1}\orcidID{0000-0002-3268-9338}
  \and
  Ornela Dardha\inst{2}\orcidID{0000-0001-9927-7875}
}
\institute{
  University of Glasgow, Scotland \\ \email{u.zalakain.1@research.gla.ac.uk}
  \and
  University of Glasgow, Scotland \\ \email{ornela.dardha@glasgow.ac.uk}
}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

@@

\begin{abstract}
  \todo{the only references here are to session types}
  \todo{fix acknowledgments}
  \todo{fix lemma, definition, theorem, example envs}
  The \picalc{} is a computational model for communication and concurrency.
  The \emph{linear} \picalc{} is a typed version of the \picalc{} where channels must be used exactly once.
  It is an underlying theoretical and practical framework on top of which more advanced types and theories can be built, such as \emph{session types} \cite{H93,THK94,HVK98}.

  We present an untyped syntax and semantics for the \picalc{}, on top of which we provide a resource-aware type system parametrised by a set of multiplicity algebras.
  This allows the user to mix \emph{linear}, \emph{graded} and \emph{shared} types, under the same unified framework.
  We use leftover typing \cite{Allais2018a} to encode our typing rules, thus avoiding the ubiquitous context splitting proofs common when dealing with linear type systems.
  We provide a framing theorem, generalise the weakening and strengthening theorems, and use them to prove subject congruence.
  We show that the type system is stable under substitution and prove subject reduction.
%
  Our formalisation is fully mechanised in Agda.

  \keywords{Pi-calculus \and Linear types \and Leftover typing \and Concurrency \and Mechanisation \and Agda}
\end{abstract}

@@

In this paper we make the following contributions:
\begin{enumerate}
  \item We use \textbf{type-level {de Bruijn indices} \cite{deBruijn1972,Dybjer1994} to introduce a syntax for \picalc{} processes that is well-scoped by construction}:
    Every free variable is accounted for in the type of the process that uses it (\autoref{syntax}).
    \todo{Do Ciccone \& Padovani also use type level de Bruijn indices?}

@@

\subsection{Summary}
\todo{shorten}
\begin{enumerate}
\item \textbf{Formalisation of the syntax and the semantics of the \picalc{}}:
 \begin{itemize}
   \item \textbf{Syntax}: we use type level {de Bruijn indices} \cite{deBruijn1972,Dybjer1994} to introduce a syntax of \picalc{} processes that is \emph{well scoped by construction}: every free variable is accounted for in the type of the process that uses it (\autoref{syntax}).
   For convenience and readability, the user can convert from and to a syntax using channel names (\autoref{from_to_deBruijn}).
   These conversion functions are shown to be inverses of one another up to $\alpha$-conversion.

@@

  \begin{equation*}
    \begin{aligned}
      \Process_n ::=& \; \PO                &&\text{(inaction)}   \\ 
      |& \; \new{} \Process_{\suc n}         &&\text{(restriction)}          \\ 
      |& \; \comp{\Process_n}{\Process_n}    &&\text{(parallel)} \\ 
      |& \; \recv{\Var_n}{}\Process_{\suc n} &&\text{(input)}                \\ 
      |& \; \send{\Var_n}{\Var_n}\Process_n  &&\text{(output)}                  
    \end{aligned}
  \end{equation*}
  \caption{Well-scoped grammar with type-level de Bruijn indices.}
  \label{fig:syntax}
\end{figure}

@@

\begin{definition}
  \label{def:unused}
  A variable $i$ is unused in $P$ ($\Unused_i \; P$) if it appears in none of its inputs and none of its outputs.
  \todo{should we have this as a figure instead?}

@@

  The type $i \type{\not\equiv} x$ unfolds to the negation of propositional equality on \textsc{Var}, i.e. $i \type{\equiv} x \to \type{\bot}$.
\end{definition}

@@

\begin{lemma}
  \label{lm:substitution-unused}
  For every variables $i$ and $j$, if $i \type{\not\equiv} j$ then $\Unused_i \allowbreak (\subst{P}{j}{i})$.
\end{lemma}
\begin{proof}
  By structural induction on \textsc{Process} and \textsc{Var}.
\end{proof}

@@

\begin{definition}[Usage algebra]
  A \emph{usage algebra} is a ternary relation $\op{x}{y}{z}$ that is \emph{partial} (as not any two multiplicities can be combined), \emph{deterministic} and \emph{cancellative} (to aid equational reasoning), \emph{associative} and \emph{commutative} (following directly from subject congruence for parallel composition), and in which the leftovers can be \emph{computed} (as to automatically update the usage context every time input and output occurs).
  It has a \emph{neutral element} $\zero$ that is absorbed on either side, and that is also \emph{minimal} (so that new resources cannot arbitrarily spring into life).
  It has an element $\one$ that is used to count inputs and outputs.
  \autoref{fig:multiplicities} defines such an algebra as a record $\Algebra_C$ on a carrier $C$.

  \begin{figure}[h]
  \begin{equation*}
  \begin{aligned}
    &\zero                   &:{} &                 &        & C \\
    &\one                    &:{} &                 &        & C \\
    &\op{\_}{\_}{\_}         &:{} &                 &        & C \to C \to C \to \Set \\
    &\func{\cdot-compute^r}  &:{} &\forall x y       & \to \; & \type{DEC} \; (\type{\exists} z  \; (\op{x}{y}{z})) \\
    &\func{\cdot-unique}     &:{} &\forall x x' y z  & \to \; & \op{x'}{y}{z} \to \op{x}{y}{z} \to x' \equiv x \\
    &\func{\cdot-unique^l}   &:{} &\forall x y y' z  & \to \; & \op{x}{y'}{z} \to \op{x}{y}{z} \to y' \equiv y \\
    &\func{\cdot-min^l}      &:{} &\forall y z        & \to \; & \op{\zero}{y}{z} \to y \equiv \zero \\
    &\func{\cdot-id^l}       &:{} &\forall x         & \to \; & \op{x}{\zero}{x} \\
    &\func{\cdot-comm}       &:{} &\forall x y z     & \to \; & \op{x}{y}{z} \to \op{x}{z}{y} \\
    &\func{\cdot-assoc}      &:{} &\forall x y z u v & \to \; & \op{x}{y}{z} \to \op{y}{u}{v} \to \\
    &                        &    &                  &        & \type{\exists} w  \; (\op{x}{u}{w} \times \op{w}{v}{z})
  \end{aligned}
  \end{equation*}
  \caption{
    Usage algebra $\Algebra_C$ on a carrier $C$.
    We use $\forall$ for universal quantification.
    The dependent product $\type{\exists}$ uses the value of its first argument in the type of its second.
    The type $\type{DEC} \; P$ is a witness of either $P$ or $P \to \type{\bot}$, where $\type{\bot}$ is the empty type with no constructors.
  }
  \label{fig:multiplicities}
  \end{figure}
\end{definition}

@@

We use indexed sets of usage algebras to allow several usage algebras to coexist in our type system with leftovers (\autoref{leftover-typing}).
\begin{definition}[Indexed set of usage algebras]
  An \emph{indexed set of usage algebras} is a type $\Idx$ of indices that is nonempty ($\type{\exists IDX}$) together with an interpretation $\Usage$ of indices into types, and an interpretation $\type{ALGEBRAS}$ of indices into usage algebras of the corresponding type (\autoref{fig:indexed-multiplicities}).
  \begin{figure}[h]
    \begin{equation*}
      \begin{aligned}
        &\Idx               &:{} &\Set \\
        &\type{\exists IDX} &:{} &\Idx \\
        &\Usage             &:{} &\Idx \to \Set \\
        &\type{ALGEBRAS}    &:{} &(idx : \Idx) \to \Algebra_{\Usage_{idx}}
      \end{aligned}
    \end{equation*}
    \caption{Indexed set of usage algebras.}
    \label{fig:indexed-multiplicities}
  \end{figure}
\end{definition}

@@

\begin{definition}[\textsc{Type} and \textsc{PreCtx}: types and typing contexts]
  A \emph{type} is either a unit type ($\unit$), a base type ($\base{n}$) or a channel type ($\channel{t}{x}$) (first two rows of \autoref{fig:prectx}).

@@

A \emph{typing context} (last two rows of \autoref{fig:prectx}) is a length-indexed list of types, and is either empty $[]$ or the result of appending a type $t$ to an existing context $\gamma$.
\end{definition}

\begin{definition}[\textsc{Ctx}: usage contexts]
  A \emph{usage context} is a context $\Ctx_{idxs}$ of pairs of usage annotations that is indexed by a length-indexed context of indices $\Idxs_n$ (\autoref{fig:ctx}).

@@

    \inferrule
        {\Gamma : \Ctx_{idxs} \\ x : \Usage_{idx} ^{\func{2}}}
        {\Gamma \comma x : \Ctx_{idxs \comma idx}}
  \end{mathpar}
  \caption{Length-indexed context of carrier indices with a context of usage annotations on top.}
  \label{fig:ctx}
  \end{figure}
\end{definition}

@@

\begin{definition}[\textsc{VarRef}: typing variable references]
  \label{def:varref}

@@

  The base case $\constr{0}$ splits the usage annotation $x$ of type $\Usage_{idx}$ into $y$ and $z$ (the leftovers).
  Note that the remaining context $\Gamma$ is preserved unused as a leftover.
  This splitting $\opsquared{x}{y}{z}$ is as per the usage algebra provided by the developer for the index $idx$.
  We use the \emph{computability} of the monoidal relation to alleviate the user from the proof burden $\opsquared{x}{y}{z}$.
  The inductive case $\suc$ appends the type $t'$ to the typing context, and the usage annotation $x'$ to both the input and output usage contexts. 
\end{definition}

@@

\begin{definition}[\textsc{Types}: typing processes]
  \label{def:types}

@@

  The inaction process in rule $\PO$ does not change usage annotations.
  The scope restriction in rule $\new$ expects three arguments: the type $t$ of data being transmitted; the usage annotation $x$ of what is being transmitted; and the multiplicity $y$ given to the channel itself.
  This multiplicity $y$ is used for both input and output, so that they are balanced.
  The continuation process $P$ is provided with the new channel with usage annotation $y \comma y$, which it must completely exhaust.
%
  The input process in rule $\recv{}{}$ requires a channel $chan_i$ at index $i$ with usage $\li$ available, such that data with type $t$ and usage $x$ can be sent over it.
  Note that the index $i$ is used in the syntax of the typed process.
  We use the leftovers $\Xi$ to type the continuation process, which is also provided with the received element --- of type $t$ and multiplicity $x$ --- at index $\constr{0}$.
  The received element $x$ must be completely exhausted by the continuation process.
%
  Similarly to input, the output process in rule $\send{}{}$ requires a channel $chan_i$ at index $i$ with usage $\lo$ available, such that data with type $t$ and usage $x$ can be sent over it.
  We use the leftover context $\Delta$ to type the transmitted data, which needs an element $loc_j$ at index $j$ with type $t$ and usage $x$, as per the type of the channel $chan_i$.
  The leftovers $\Xi$ are used to type the continuation process.
  Note that both indices $i$ and $j$ are used in the syntax of the typed process.
%
  Parallel composition in rule $\comp{}{}$ uses the leftovers of the left-hand process to type the right-hand process.
  By keeping track in the typing derivation of $P$ of the resources $P$ uses, we can use them to type $Q$ and save the user from having to provide a top-down proof of context split.
\end{definition}

@@

\begin{theorem}[Framing]
  \label{thm:framing}
  Let $P$ be well typed in $\types{\gamma}{\Gamma_l}{P}{\Xi_l}$.
  Let $\Delta$ be such that $\opctx{\Gamma_l}{\Delta}{\Xi_l}$.
  Let $\Gamma_r$ and $\Xi_r$ be arbitrary contexts such that $\opctx{\Gamma_r}{\Delta}{\Xi_r}$.
  Then $\types{\gamma}{\Gamma_r}{P}{\Xi_r}$.
\end{theorem}

\paragraph*{Weakening}
This property states that inserting a new variable into the context preserves the well-typedness of a process as long as the usage annotation of the inserted variable is preserved as a leftover.
Let $\func{ins}_i$ insert an element into a context at position $i$ --- for simplicity, we use it both to insert types into typing contexts and usage annotations into usage contexts.
\begin{theorem}[Weakening]
  \label{thm:weakening}
  Let $P$ be well typed in $\types{\gamma}{\Gamma}{P}{\Xi}$.
  Then, lifting every variable greater than or equal to $i$ in $P$ is well typed in
  $\types{\func{ins}_i \; t \; \gamma}{\func{ins}_i \; x \; \Gamma}{\func{lift}_i \; P}{\func{ins}_i \; x \; \Xi}$.
\end{theorem}

\paragraph*{Strengthening}
This property states that removing an unused variable preserves the well-typedness of a process.
Let $\func{del}_i$ delete the element at position $i$ from a context --- for simplicity, we use it both to delete types from typing contexts and usage annotations from usage contexts.
\begin{theorem}[Strengthening]
  \label{thm:strengthening} 
  Let $P$ be well typed in $\types{\gamma}{\Gamma}{P}{\Xi}$.
  Let $i$ be a variable not in $P$, such that $uP \; : \; \Unused_i \; P$.
  Then lowering every variable greater than $i$ in $P$ is well typed in $\types{\func{del}_i \; \gamma}{\func{del}_i \; \Gamma}{\func{lower}_i \; P \; uP}{\func{del}_i \; \Xi}$.
\end{theorem}

\paragraph*{Exchange}
This property states that the exchange of two variables preserves the well-typedness of a process.
We extend $\func{exchange}_i$  introduced in \autoref{structural-congruence} to exchange types in typing contexts and usage annotations in usage contexts.
\begin{theorem}[Exchange]
  \label{thm:exchange}
  Let $P$ be well typed in $\types{\gamma}{\Gamma}{P}{\Xi}$.
  Then, $\types{\func{exchange}_i \; \gamma}{\func{exchange}_i \; \Gamma}{\allowbreak \func{exchange}_i \; P}{\func{exchange}_i \; \Xi}$.
\end{theorem}

@@

\begin{lemma}
  \label{lm:lower-lift}
  The function $\func{lower}_i \; P \; uP$ has an inverse $\func{lift}_i \; P$ that increments every $\textsc{Var}$ greater than or equal to $i$, such that $\func{lift}_i \; (\func{lower}_i \; P \; uP) \equiv P$.
\end{lemma}
\begin{proof}
  By structural induction on \textsc{Process} and \textsc{Var}.
\end{proof}

\begin{lemma}
  \label{lm:exchange-exchange}
  The function $\func{exchange}_i \; P$ is its own inverse: $\func{exchange}_i \; (\func{exchange}_i \; P) \equiv P$.
\end{lemma}
\begin{proof}
  By structural induction on \textsc{Process} and \textsc{Var}.
\end{proof}

\begin{lemma}
  \label{lm:types-unused}
  For all well-typed processes $\types{\gamma}{\Gamma}{P}{\Xi}$, if the variable $i$ is unused within $P$, then $\Gamma$ at $i$ is equivalent to $\Xi$ at $i$.
\end{lemma}
\begin{proof}
  By induction on \textsc{Process} and \textsc{Var}.
\end{proof}

@@

\begin{theorem}[Subject congruence]
  \label{thm:subject-congruence}
  If $P \eq{} Q$ and $\types{\gamma}{\Gamma}{P}{\Xi}$, then $\types{\gamma}{\Gamma}{Q}{\Xi}$.
\end{theorem}
